% Control Structures
% Biostatistics 140.776
%

```{r, include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

# Control Structures


Control structures in R allow you to control the flow of execution of a series of R expressions. Commonly used control structures are

- `if` and `else`: testing a condition and acting on it

- `for`: execute a loop a fixed number of times 

- `break`: break the execution of a loop

- `next`: skip an interation of a loop

Most control structures are not used in interactive sessions, but rather when writing functions or longer expresisons.

# `if`

Test a condition and act on it depending on whether it's true or false. 

```r
if(<condition>) {
        ## do something
} 
## Continue with rest of code
```


# `if`-`else`

The previous code does nothing if the condition is false. If you have an action you want to execute when the condition is false, then you need an `else` clause.

```r
if(<condition>) {
        ## do something
} 
else {
        ## do something else
}
```

# `if`-`else`

You can have a series of tests by following the initial `if` with any number of `else if`s.

```r
if(<condition1>) {
        ## do something
} else if(<condition2>)  {
        ## do something different
} else {
        ## do something different
}
```

# Alternate `if`-else`

Here is an example of a valid if/else structure.

```{r}
## Generate a uniform random number
x <- runif(1, 0, 10)  
if(x > 3) {
        y <- 10
} else {
        y <- 0
}
```

The value of `y` is set depending on whether `x > 3` or not. 

# `if`-`else`

You could have a series of if clauses that always get executed if their respective conditions are true.

```r
if(<condition1>) {

}

if(<condition2>) {

}
```


# `for` Loops

For loops take an **interator variable** and assign it successive values from a **sequence** or **vector**. 

* Most commonly used for iterating over the elements of an object (list, vector, etc.)

```{r, eval=FALSE}
numbers <- rnorm(5)
for(i in 1:5) {
        print(numbers[i])
}
```

This loop takes the `i` variable and in each iteration of the loop gives it values 1, 2, 3, 4, 5, executes the code within the curly braces, and then the loop exits.


# `for` Loops

There are many ways to specify a `for` loop.

```{r}
x <- c("a", "b", "c", "d")

for(i in 1:4) {
        ## Print out each element of 'x'
        print(x[i])  
}
```

# `for` Loops

The `seq_along()` function is commonly used in conjunction with for loops in order to generate an integer sequence based on the length of an object (in this case, the object `x`).

```{r}
x <- c("a", "b", "c", "d")

## Generate a sequence based on length of 'x'
for(i in seq_along(x)) {   
        print(x[i])
}
```


# `for` Loops

It is not necessary to use an index-type variable. 

```{r}
x <- c("a", "b", "c", "d")

for(letter in x) {
        print(letter)
}
```

# `for` Loops

For one line loops, the curly braces are not strictly necessary.

```{r}
x <- c("a", "b", "c", "d")
for(i in 1:4) 
        print(x[i])
```

# `for` Loops

Or simply 

```{r}
x <- c("a", "b", "c", "d")
for(i in 1:4) print(x[i])
```

Curly braces are sometimes useful even for one-line loops, because that way if you decide to expand the loop to multiple lines, you won't be burned because you forgot to add curly braces (and you *will* be burned by this).


# Nested `for` loops

`for` loops can be nested inside of each other.

```{r,eval=FALSE}
x <- matrix(1:6, 2, 3)

for(i in seq_len(nrow(x))) {
        for(j in seq_len(ncol(x))) {
                print(x[i, j])
        }   
}
```

* Nested loops are commonly needed for multidimensional or hierarchical data structures (e.g. matrices, lists). 

* Nesting beyond 2 to 3 levels often makes it difficult to read or understand the code. 

* If you find yourself in need of a large number of nested loops, you may want to break up the loops by using functions


# `next`, `break`

`next` is used to skip an iteration of a loop. 

```{r,eval=FALSE}
for(i in 1:100) {
        if(i <= 20) {
                ## Skip the first 20 iterations
                next                 
        }
        ## Do something here
}
```


# `next`, `break`

`break` is used to exit a loop immediately, regardless of what
iteration the loop may be on.

```{r,eval=FALSE}
for(i in 1:100) {
      print(i)

      if(i > 20) {
              ## Stop loop after 20 iterations
              break  
      }		
}
```


# Summary

- Control structures like `if`-`else` and `for` allow you to
  control the flow of an R program

- Control structures mentiond here are primarily useful for writing programs; for command-line interactive work, the "apply" functions are typically more useful.

