% Dates and Times in R
% Biostatistics 140.776
%

```{r, include=FALSE}
options(width = 50)
knitr::opts_chunk$set(comment = NA)
```

# Dates and Times in R

R has developed a special representation of dates and times

- Dates are represented by the `Date` class

- Times are represented by the `POSIXct` or the `POSIXlt` class

- Dates are stored internally as the number of days since 1970-01-01

- Times are stored internally as the number of seconds since
  1970-01-01

# The `lubridate` package

* The `lubridate` package is a very useful package for dealing with all the little annoying aspects of dates/times

* Largely replaces the default date/time functions in base R

* Methods for date/time arithmetic

* Handles time zones, leap year, leap seconds, etc.


# Dates in R

Dates are represented by the `Date` class and can be coerced from a character string using the `ymd()` function.

```{r,message=FALSE}
library(lubridate)
x <- ymd("1970-01-01")
x
```

`Date` objects have their own special print methods that will always format as "YYYY-MM-DD".

# Alternate Formulations

Different locales have different ways formatting dates

```{r}
ymd("2016-09-13")  ## International standard
mdy("09-13-2016")  ## Mostly U.S.
dmy("13-09-2016")  ## Europe
```
All of the above are valid and lead to the exact same object.

# Times in R

Times are represented using the `POSIXct` or the `POSIXlt` class

- `POSIXct` is just a very large integer under the hood; it is a useful class when you want to store times in something like a data frame

- `POSIXlt` is a list underneath and it stores a bunch of other useful information like the day of the week, day of the year, month, day of the month

There are a number of generic functions that work on dates and times

- `weekdays`: give the day of the week

- `month`: give the month name (possibly abbreviated)

- `quarter`: give the quarter number (1, 2, 3, 4)



# Times in R

Times can be coerced from a character string with `ymd_hms()`

```{r}
ymd_hms("2016-09-13 14:00:00")
ymd_hms("2016-09-13 14:00:00", tz = "America/New_York")
ymd_hms("2016-09-13 14:00:00", tz = "")
```

# Time Zones!

Time zones were created to make your data analyses more difficult.

* `ymd_hms()` function will by default use UTC as the time zone

* Specifying `tz = ""` will use the local time zone

* Better to specify time zone when possible to avoid ambiguity

You can go to Wikipedia to find the list of time zones [http://goo.gl/xJ8K6q]()



# Specifying Times in R

Finally, there is the `strptime()` function in case your dates are
written in a different format

```{r,tidy=FALSE}
datestring <- c("January 10, 2012 10:40", 
                "December 9, 2011 9:10")
x <- strptime(datestring, "%B %d, %Y %H:%M", 
              tz = "America/Los_Angeles")
x
```

* Check `?strptime` for details of formatting strings

* When reading in data with `read_csv()`, you may need to read in as character first and then convert to date/time


# Operations on Dates and Times

You can add and subtract dates and times. You can do comparisons too (i.e. `==`, `<=`)

```{r, error=TRUE}
x <- ymd("2012-01-01", tz = "")  ## Midnight
y <- dmy_hms("9 Jan 2011 11:34:21", tz = "")
x - y
x + y  ## Nope!
```


# Operations on Dates and Times

```{r}
y + 1  ## Add a number to a time
y <- date(y)  ## Just keep the date portion
y
y + 1  ## Add a number to a date
```

# Operations on Dates and Times

Even keeps track of leap years, leap seconds, daylight savings, and time zones.

```{r}
x <- ymd("2012-03-01")
y <- ymd("2012-02-28")
x - y
x <- ymd_hms("2012-10-25 01:00:00", tz = "")
y <- ymd_hms("2012-10-25 06:00:00", tz = "GMT")
y - x
```

# Extracting Elements of Dates/Times

There are a set of helper functions in `lubridate` that can extract sub-elements of dates/times

```{r}
x <- ymd_hms(c("2012-10-25 01:13:46",
               "2015-04-23 15:11:23"), tz = "")
year(x)
month(x)
day(x)
weekdays(x)
```


# Extracting Elements of Dates/Times

```{r}
minute(x)
second(x)
hour(x)
week(x)
```

# Summary

- Dates and times have special classes in R that allow for numerical and statistical calculations

- Dates use the `Date` class

- Times use the `POSIXct` and `POSIXlt` class

- Character strings can be coerced to Date/Time classes using the `ymd()` and `ymd_hms()` functions. In strange cases, you can use the `strptime()`or the `as.Date()` functions.

- The `lubridate` package is essential for manipulating date/time data

