% Managing Data with `dplyr` and `tidyr`
% Biostatistics 140.776
%

```{r, include=FALSE}
options(width = 50)
knitr::opts_chunk$set(comment = NA)
```

# dplyr

The data frame is a key data structure in statistics and in R.

* There is one observation per row

* Each column represents a variable or measure or characteristic

* Primary implementation that you will use is the default R
  implementation

* Other implementations, particularly relational databases systems


# dplyr

* Developed by Hadley Wickham of RStudio

* An optimized and distilled version of `plyr` package (also by Hadley)

* Does not provide any "new" functionality per se, but **greatly**
  simplifies existing functionality in R

* Provides a "grammar" (in particular, verbs) for data manipulation

* Is **very** fast, as many key operations are coded in C++


# dplyr Verbs

* `select`: return a subset of the columns of a data frame

* `filter`: extract a subset of rows from a data frame based on
  logical conditions

* `arrange`: reorder rows of a data frame


* `rename`: rename variables in a data frame

* `mutate`: add new variables/columns or transform existing variables

* `summarise` / `summarize`: generate summary statistics of different
  variables in the data frame, possibly within strata

There is also a handy `print` method that prevents you from printing a
lot of data to the console.



# dplyr Properties

* The first argument is a data frame.

* The subsequent arguments describe what to do with it, and you can
  refer to columns in the data frame directly without using the $
  operator (just use the names).

* The result is a new data frame

* Data frames must be properly formatted and annotated for this to all
  be useful


# Load the `dplyr` package


This step is important!

```{r}
library(dplyr)
```


# `select`

```{r}
chicago <- readRDS("chicago.rds")
dim(chicago)
head(select(chicago, 1:5))
```


# `select`

```{r}
names(chicago)[1:3]
head(select(chicago, city:dptp))
```

# `select`

In dplyr you can do

```{r,eval=FALSE}
head(select(chicago, -(city:dptp)))
```

Equivalent base R

```{r,eval=FALSE}
i <- match("city", names(chicago))
j <- match("dptp", names(chicago))
head(chicago[, -(i:j)])
```



# `filter`

```{r}
chic.f <- filter(chicago, pm25tmean2 > 30)
head(select(chic.f, 1:3, pm25tmean2), 10)
```

# `filter`

```{r}
chic.f <- filter(chicago, pm25tmean2 > 30 & tmpd > 80)
head(select(chic.f, 1:3, pm25tmean2, tmpd), 10)
```


# `arrange`

Reordering rows of a data frame (while preserving corresponding order
of other columns) is normally a pain to do in R.

```{r}
chicago <- arrange(chicago, date)
head(select(chicago, date, pm25tmean2), 3)
tail(select(chicago, date, pm25tmean2), 3)
```

# `arrange`

Columns can be arranged in descending order too.

```{r}
chicago <- arrange(chicago, desc(date))
head(select(chicago, date, pm25tmean2), 3)
tail(select(chicago, date, pm25tmean2), 3)
```


# `rename`

Renaming a variable in a data frame in R is surprising hard to do!

```{r,tidy=FALSE}
head(chicago[, 1:5], 3)
chicago <- rename(chicago, dewpoint = dptp, 
                  pm25 = pm25tmean2)
head(chicago[, 1:5], 3)
```


# `mutate`

```{r, tidy=FALSE}
chicago <- mutate(chicago, 
                  pm25detrend=pm25-mean(pm25, na.rm=TRUE))
head(select(chicago, pm25, pm25detrend))
```

# `group_by`

Generating summary statistics by stratum

```{r, tidy=FALSE}
chicago <- mutate(chicago, 
                  tempcat = factor(tmpd > 90, 
                                   labels = c("cold", "hot")))
hotcold <- group_by(chicago, tempcat)
summarize(hotcold, pm25 = mean(pm25, na.rm = TRUE), 
          o3 = max(o3tmean2, na.rm = TRUE), 
          no2 = median(no2tmean2, na.rm = TRUE))
```


# `group_by`

Generating summary statistics by stratum

```{r, tidy=FALSE}
chicago <- mutate(chicago, 
                  year = as.POSIXlt(date)$year + 1900)
years <- group_by(chicago, year)
summarize(years, pm25 = mean(pm25, na.rm = TRUE), 
          o3 = max(o3tmean2, na.rm = TRUE), 
          no2 = median(no2tmean2, na.rm = TRUE))
```

```{r,echo=FALSE}
chicago$year <- NULL  ## Can't use mutate to create an existing variable
```


# `%>%`

```{r,tidy=FALSE,eval=FALSE}
chicago %>% mutate(year = as.POSIXlt(date)$year + 1900) 
	%>% group_by(year) 
	%>% summarize(pm25 = mean(pm25, na.rm = TRUE), 
          o3 = max(o3tmean2, na.rm = TRUE), 
          no2 = median(no2tmean2, na.rm = TRUE))
```

```{r,echo=FALSE}
chicago %>% mutate(year = as.POSIXlt(date)$year + 1900) %>% group_by(year) %>% 
summarize(pm25 = mean(pm25, na.rm = TRUE), o3 = max(o3tmean2, na.rm = TRUE), no2 = median(no2tmean2, na.rm = TRUE))

```


# dplyr

Once you learn the dplyr "grammar" there are a few additional benefits

* dplyr can work with other data frame "backends"

* `data.table` for large fast tables

* SQL interface for relational databases via the DBI package

# tidyr

The `tidyr` package helps with manipulation of data frames between "wide" and "long" formats, depending on what you're trying to do.

* Sometimes the meaning of a "variable" depends on the application

* Sometimes PM10, O3, NO2 are all different variables with continuous levels

* Sometimes "Pollutant" is the variable with levels "PM10", "O3", and "NO2"

# Long Format

```{r, include=FALSE}
library(tidyr)
chicago <- readRDS("chicago.rds")
chicago <- select(chicago, date, pm10tmean2, o3tmean2, no2tmean2) %>%
        rename(pm10 = pm10tmean2, o3 = o3tmean2, no2 = no2tmean2) %>%
        gather(pollutant, level, -date) %>%
        arrange(date, pollutant)
```

Here are the Chicago pollution data in long format

```{r}
head(chicago)
```

# Long Format

```{r}
stats <- group_by(chicago, pollutant) %>% 
        summarize(mean = mean(level, na.rm = TRUE),
                  median = median(level, na.rm = TRUE),
                  max = max(level, na.rm = TRUE))
stats
```

# `gather`

An alternate representation could have three variables: pollutant, statistic, and value

* `gather` is a function that "gathers" multiple columns and essential sticks them into one column

* The names of multiple columns become levels of a single variable

* In this case mean, median, max --> levels of a "statistic" variable

# `gather`

```{r}
gather(stats, statistic, value, -pollutant)
```

# `spread`

The `spread` function does the inverse of the `gather` function

* `spread` takes a single variable (with multiple levels) and *spreads* them across multiple columns

* Sometimes more intuitive if you want to compute a statistic across multiple levels/variables

* e.g. Compute the maximum of three different pollutants on each day and create a new variable

# `spread`

```{r}
wide <- spread(chicago, pollutant, level) 
head(wide)
```

# `spread`

```{r}
mutate(wide, max = pmax(no2, o3, pm10, na.rm = TRUE)) %>%
        head
```

# `separate`

Sometimes you need to split one column into two separate columns.

```{r, include=FALSE}
x <- paste(rep(c("happy", "sad"), each = 3), rep(c("working", "partying"), 3), sep = "-")
m <- data.frame(mood = x, day = c(rep("Sunday", 4), rep("Monday", 2)))
```

```{r}
m
```

# `separate`

```{r}
m <- separate(m, mood, c("state", "activity"), sep = "-")
m
```

# `unite`

The inverse of `separate`

```{r}
unite(m, mood, state, activity, sep = "-")
```

# Summary

`dplyr`

* Verbs/functions for manipulating data frames in tidy format

* select, filter, arrange, group_by, summarize, rename, mutate

`tidyr`

* Transform data frames from wide to long formats

* spread, gather, separate, unite