% Functional Programming
% Biostatistics 140.776
%

```{r,include=FALSE}
knitr::opts_chunk$set(comment = NA)
```

# What is Functional Programming?

Functional programming concentrates on four constructs:

1. Data (numbers, strings, etc)
2. Variables (function arguments)
3. Functions
4. Function Applications (evaluating functions given arguments and/or data)


# What is Functional Programming?

By now you're used to treating variables inside of functions as data:

* numbers and strings
* data structures like lists and vectors. 

With functional programming you can also

* provide a **function** as an argument to another function
* return a function as the result of a function

The `sapply()` function is a an example of supplying a function as an argument to another function

# Simple Example

The following function returns a function as its "result".

```{r}
adder_maker <- function(n) {
        function(x){
                n + x
        }
}
```

This function "constructs" functions based on its argument `n`.

# Simple Example

```{r}
## Make a function that adds 3 to its argument
add3 <- adder_maker(3)
add3(5)

## Make a function that adds 2 to its argument
add2 <- adder_maker(2)
add2(5)
```

# Simple Example

A few notes on this example:

* This example works because each of the `add` functions "knows" its value of `n`

* Lexical scoping rules mean that both `add2()` and `add3()` look up the value of `n` in the environment where they were defined

* Each function was defined inside the `adder_maker()` function which supplied the value of `n` via the user


# Functional Programming


The functional programming approach essentially allows you to **write code** based on **data** supplied by the user!

1. Usual model: Data --> Code --> "Results"

2. Functional model: Data --> Code --> Code (!)

It's a very powerful technique that can allow you to develop so-called "intelligent" systems


# The `purrr` Package

* There are groups of functions that are essential for functional programming: map, reduce, filter, compose, search

* In most cases they take a function and a data structure as arguments, and that function is applied to that data structure in some way. 

* The `purrr` package contains many of these functions 

* Functional programming is concerned mostly with lists and vectors (for data structures)

# Map

A key function in the `purrr` package is the `map()` function, which has two arguments:

* `.x` A data structure like a list or vector

* `.f` A function to be applied, a formula (converted into a function), or an atomic vector

`map()` always returns a list, regardless of the input. It works much like `lapply()`. 

# Map

```{r}
library(purrr)
num2word <- function(x) {
        words <- c("one", "two", "three", "four", "five")
        words[x]
}
map(3:1, num2word)
```


# Map with Simplification

```{r}
map_chr(5:1, num2word)
```

# Map

Often we want to map a function to an object (list/vector) and then extract an element of each of the return values.

This code splits the MIE data into separate rooms, fits a linear model to each subset, and then extracts the coefficients.

```{r,message=FALSE}
results <- split(airquality, airquality$Month) %>%
        map(~ lm(Ozone ~ Temp, data = .x)) %>%
        map("coefficients")
```

# Map

```{r}
results[1:3]
```


# Map

Using `map()` with a formula argument "auto-generates" an anonymous function, which is more compact than using something like `lapply()`. 

```{r,eval=FALSE}
split(airquality, airquality$Month) %>%
        map(~ lm(Ozone ~ Temp, data = .x))
```

is equivalent to

```{r,eval=FALSE}
lapply(split(airquality, airquality$Month), 
       function(.x) {
               lm(Ozone ~ Temp, data = .x)
       })
```

The version using `map()` is more readable, modular, and highlights what is happening in the right sequence.


# Conditional Map

The `map_if()` function takes a **predicate** and then a function for application. A predicate is a function that returns TRUE or FALSE given some input.


```{r}
## A predicate function
is_even <- function(x) {
        x %% 2 == 0
}
square <- function(y){
        y^2
}
map_if(1:5, is_even, square) %>% unlist
```

If the predicate evalutes to FALSE, the argument just passes through.


# Reduce

List or vector reduction is done with the `reduce()` function. It is a kind of looping that

1. Combines the first element of a vector with the second element of a vector; 

2. then that combined result is combined with the third element of the vector;

3. and so on until the end of the vector is reached. 

The function to be applied should take at least **two** arguments.

# Reduce

Often, after mapping a function to a list/vector, you will want to combine the results at the end.

```{r}
map_if(1:5, is_even, square) %>%
        reduce(function(x, y) {
                x + y
        })
```

This is overkill since the result is just a simple list; we could have just used `sum()`.

But what if the `map()` function returns a more complicated object?

# Reduce

```{r}
split(airquality, airquality$Month) %>%
        map(~ lm(Ozone ~ Temp, data = .x)) %>%
        map("coefficients") %>%
        reduce(function(x, y) {
                ## 'x' and 'y' are numeric vectors
                cbind(x, y)
        })

```


# Search

You can search for specific elements of a vector using the

* `contains()`: will return `TRUE` if a specified element is present in a vector, otherwise it returns `FALSE` 

* `detect()`: takes a vector and a predicate function as arguments and it returns the first element of the vector for which the predicate function returns `TRUE

# Search

```{r}
purrr::contains(letters, "a")
purrr::contains(letters, "A")
```

# Search

```{r}
detect(20:40, function(x){
  x > 22 && x %% 2 == 0
})
```


# Filter

The various filter functions are

* `keep()`: Keep elements that satisfy a predicate

* `discard()`: Remove elements that satisfy a predicate

* `every()`: returns TRUE only if every element in the vector satisfies the predicate function

* `some()`: returns TRUE if at least one element in the vector satisfies the predicate function


# Filtering

```{r}
keep(1:20, function(x){
  x %% 2 == 0
})
```

```{r}
discard(1:20, function(x){
  x %% 2 == 0
})
```


# Filtering

```{r}
every(1:20, function(x){
  x %% 2 == 0
})
```

```{r}
some(1:20, function(x){
  x %% 2 == 0
})
```



# Compose

The `compose()` function take any number of functions and combines them into a single function. 

```{r}
nlevels <- compose(length, unique)
nlevels(c("a","a", "a", "b", "b"))
```

This is equivalent to

```{r}
nlevels <- function(x) {
        length(unique(x))
}
```

The evaluation of functions is done from right to left.

# Compose

You can also do more unusual things like

```{r}
not_null <- compose("!", is.null)
not_null(NULL)
not_null(4)
```

# Compose

You can also build upon little building blocks

```{r}
add1 <- function(x) {
        x + 1
}
add3 <- compose(add1, add1, add1)
add3(8)
```

# Compose

When fitting linear models, I do this a lot:

```{r}
summary_lm <- compose(summary, lm)
summary_lm(Ozone ~ Temp, data = airquality)
```


# Compose

Or even

```{r}
coef_lm <- compose(coef, lm)
coef_lm(Ozone ~ Temp, data = airquality)
```


# Summary

* Functional programming treats functions as "first class" citizens, along with data, and parameters

* Building functions that adapt to data is a key element

* Map, reduce, filter, search, and compose are key concepts

* Often can be a clearer way to express code (if not more succinct or faster)

* Functional programming often extends easily to parallel programming

















