% Regular Expressions in R
% Biostatistics 140.776
%

```{r, include=FALSE}
options(width = 50)
knitr::opts_chunk$set(comment = NA)
```


# Regular Expression Functions

The primary R functions for dealing with regular expressions are

* `grep`, `grepl`: Search for matches of a regular
  expression/pattern in a character vector; either return the indices
  into the character vector that match, the strings that happen to
  match, or a TRUE/FALSE vector indicating which elements match

* `regexpr`, `gregexpr: Search a character vector for regular
  expression matches and return the indices of the string where the
  match begins and the length of the match

* `sub`, `gsub`: Search a character vector for regular
  expression matches and replace that match with another string

* `regexec`: Easier to explain through demonstration.

# grep

```{r loaddata,cache=TRUE}
library(readr)
commits <- read_lines("commit_logs_strip.txt.bz2")
head(commits)
```

# grep

How many commits are there?

```{r grep1,cache=TRUE}
g <- grep("^commit", commits)
head(g)
length(g)
```

# grep

Sometimes you want `grep()` to return the value instead of the index

```{r grep2,cache=TRUE}
g <- grep("^commit", commits, value = TRUE)
head(g)
```

# grep

Who are the authors of these commits?

```{r,cache=TRUE,warning=FALSE}
g <- grep("^Author", commits, value = TRUE, perl = TRUE)
head(g)
length(unique(g))
```


# grep, grepl

By default, `grep()` returns the *indices* into the character vector where the regex pattern matches.

```{r}
head(state.name)
grep("^New", state.name)
```

`grepl()` returns a logical vector indicating which element matches.

```{r}
i <- grepl("^New", state.name)
head(i, 10)
```

# grep

Some limitations of `grep()`:

* The `grep()` function tells you which strings in a character vector match a certain pattern but it doesn’t tell you exactly where the match occurs or what the match is (for a more complicated regex).

* The `regexpr()` function gives you the index into each string where the match begins and the length of the match for that string.

* `regexpr()` only gives you the first match of the string (reading left to right). `gregexpr()` will give you all of the matches in a given string.

# regexpr

How can we obtain the email addresses of the authors?

```{r,cache=TRUE}
commits[12:24]
```

What if we use the regex `<(.*)>` and search for that?

# regexpr

We need to search the Author line for a pattern. We can first `grep` the Author lines and then search for a pattern.

```{r,cache=TRUE}
author <- grep("^Author:", commits, value = TRUE)
head(author, 3)
r <- regexpr("<.*>", author)
str(r)
```

# regexpr

* `regexpr()` returns a vector of integers indicating where the match starts

* The attribute `match.length` indicates how long the match is

* If there's no match, `regexpr()` returns `-1` with a `match.length` of `-1`.

The obvious way to select out a match is to use the indices and the `substr()` function.

```{r}
substr(author[1], 15, 15 + 6 - 1)
substr(author[3], 14, 14 + 6 - 1)
```


# regmatches

We can also use the `regmatches()` function to just grab all of the matches at once.

```{r,cache=TRUE}
r <- regexpr("<.*>", author)
m <- regmatches(author, r)
head(m)
```


# sub/gsub

But we still don't have actual email addresses. We need to remove the `<` and `>` characters. We can use the `sub()` function for that.

```{r}
sub("<", "", m[1:5])
sub(">", "", m[1:5])
```
But we want to remove *both* `<` and `>`!

# sub/gsub

We can use a regular expression in `sub()`.

```{r}
sub("<|>", "", m[1:5])
```

`gsub()` substitutes all occurrences of the regex (`g` is for "global").

```{r}
gsub("<|>", "", m[1:5])
```


# regexec

The `regexec()` function can make the previous task a bit simpler by using *parenthesized sub-expressions*.

```{r}
author[1]
```

We can capture the email address portion of the line with parentheses.

```{r}
regexec("^Author: [^ ]+ <(.*)>", author[1])
```

# regexec

```{r}
r <- regexec("^Author: [^ ]+ <(.*)>", author[1])
regmatches(author[1], r)
```

# regexec

When were all of the commits made?

```{r,cache=TRUE}
r <- regexec("^Date: +(.*)$", commits)
m <- regmatches(commits, r)
u <- sapply(m, length) > 0
dates <- sapply(m[u], function(x) x[2])
dates <- strptime(dates, "%a %b %d %H:%M:%S %Y", 
                  tz = "America/New_York")
```

# Histogram

You can make a histogram of the dates

```{r}
hist(dates, "month", freq = TRUE)
```

# Summary

The primary R functions for dealing with regular expressions are

* `grep`, `grepl`: Search for matches of a regular expression/pattern in a character vector

* `regexpr`, `gregexpr`: Search a character vector for regular expression matches and return the indices where the match begins; useful in conjunction with regmatches

* `sub`, `gsub`: Search a character vector for regular expression matches and replace that match with another string

* `regexec`: Gives you indices of parethensized sub-expressions.

















