---
title: "Homework 2 Solution"
output: 
  html_document: 
    keep_md: yes
permalink: /homework/
layout: page
---

```{r,echo=FALSE}
knitr::opts_chunk$set(comment = NA, fig.path = "homework/")
```

## Homework 2: Functions


### Part 1

Write a function that computes the factorial of an integer greater than or equal to 0. The function should have the following skeleton.

```{r}
Factorial <- function(n) {
        if(n == 0) 
                return(1)
        else
                n * Factorial(n - 1)
}
```

```{r}
Factorial(5)
```

```{r}
Factorial2 <- function(n) {
        if(n == 0)
                return(1)
        vals <- seq(n, 1)
        Reduce("*", vals)
}

Factorial3 <- function(n) {
        if(n == 0)
                return(1)
        r <- 1
        for(val in seq_len(n)) {
                r <- r * val
        }
        r
}

Factorial4 <- function(n) {
        if(n == 0)
                return(1)
        vals <- seq(n, 1)
        prod(vals)
}
```

The `microbenchmark` package allows us to time the various implementations of the factorial function, including R's built in version.

```{r}
library(ggplot2)
library(microbenchmark)
m <- microbenchmark(Factorial(5), Factorial2(5), 
                    Factorial3(5), Factorial4(5),
                    factorial(5), times = 200L)
m
autoplot(m)
```


### Part 2

The data for this part come from a study of indoor air pollution and respiratory disease conducted here at Johns Hopkins. A high-resolution air pollution monitor was placed in each home to collect continuous levels of particulate matter over the period of a few days (each measurement represents a 5-minute average). In addition, measurements were taken in different rooms of the house as well as on multiple visits. 

```{r,message=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
mie <- read_csv("data/MIE.csv", col_types = "iicdiDc")
```


```{r,message=FALSE}
library(dplyr)
viz_mie <- function(data, idviz) {
        idset <- unique(data$id)
        if(!(idviz %in% idset))
                stop("id ", idviz, " is not in the dataset")
        p <- ggplot(data = filter(data, id == idviz)) +
                geom_line(aes(x = timepoint, y = value)) + 
                facet_grid(room ~ visit)
        print(p)
}
```

```{r}
viz_mie(mie, 14)
viz_mie(mie, 54)
```


```{r}
viz_mie <- function(data, visitnum) {
        p <- ggplot(data = filter(data, visit == visitnum)) +
                geom_line(aes(x = timepoint, y = value)) + 
                facet_grid(id ~ room)
        print(p)
}
```
```{r}
viz_mie(mie, 0)
```

### Part 3

The `bootstrap()` function computes the 95% confidence interval for the median of a vector of numbers using the bootstrap procedure and the percentile method. 

The function takes as input `x` which should be a numeric vector. If it is not numeric, an attempt is made to coerce the vector to be numeric. Missing values in `x` are allowed and are ignored when computing the median.

The second argument is `N` which indicates the number of bootstrap iterations. The default is 1000, which is the minimum required for computing 95% confidence intervals. If you specify a value of `N` less than 1000, a warning is given.

Finally, the `seed` argument provides a value for the random number generator seed to ensure reproducible computations. This should be a numeric value. The default `NULL` just uses whatever the current value is (and hence is not reproducible).


```{r}
bootstrap <- function(x, N = 1000, seed = NULL) {
        if(!is.vector(x))
                stop("'x' should be a vector")
        if(!is.numeric(x))
                x <- as.numeric(x)
        if(N < 1000)
                warning("percentile intervals require N >= 1000")
        use <- !is.na(x)
        if(any(!use)) {
                message("removing missing values")
                x <- x[use]
        }
        set.seed(seed)
        b <- replicate(N, {
                x.new <- sample(x, replace = TRUE)
                median(x.new)
        })
        quantile(b, c(0.025, 0.975))
}
```

```{r}
source("data/median_testdata.R")
bootstrap(x1, seed = 10)
bootstrap(x2, seed = 10)
bootstrap(x3, seed = 10)
bootstrap(x4, seed = 10)
bootstrap(x5, seed = 10)
```

