---
title: "Homework 1 Solution"
output: 
  html_document: 
    keep_md: yes
layout: page
---

```{r,include=FALSE}
knitr::opts_chunk$set(comment = NA)
```


## Homework 1: Tidy Data

### Part 1

Load the `WorldPhones` dataset in the `datasets` package with


```{r}
library(datasets)
data(WorldPhones)
WorldPhones
```

This dataset gives the number of telephones in various regions of the world (in thousands). The regions are: North America, Europe, Asia, South America, Oceania, Africa, Central America and data are available for the years 1951, 1956, 1957, 1958, 1959, 1960, 1961.

Use the functions in `dplyr` and `tidyr` to produce a data frame that looks like this.

```{r, message=FALSE}
library(dplyr)
library(tidyr)
tbl_df(WorldPhones) %>% 
        mutate(year = row.names(WorldPhones)) %>%
        gather(country, number, -year)
```

You may need to use functions outside these packages to obtain this result.

### Part 2

```{r}
library(readr)
spec <- read_csv("data/SPEC_2014.csv.bz2", progress = FALSE)
names(spec)
```

Use the functions in the `dplyr` package to answer the following questions:

1. What is average value of "Bromine PM2.5 LC" in the state of Wisconsin in this dataset?

```{r}
filter(spec, Parameter.Name == "Bromine PM2.5 LC" & State.Name == "Wisconsin") %>%
        summarize(avg = mean(Sample.Value))
```

2. Calculate the average of each chemical constituent across all states/monitors and all time points. Which constituent has the highest average level?

```{r}
group_by(spec, Parameter.Name) %>% 
        summarize(avg = mean(Sample.Value)) %>%
        arrange(desc(avg))                          
```

3. Which monitoring site has the highest levels of "Sulfate PM2.5 LC"? Indicate the state code, county code, and site number.

```{r}
filter(spec, Parameter.Name == "Sulfate PM2.5 LC") %>%
        group_by(State.Code, County.Code, Site.Num) %>%
        summarize(avg = mean(Sample.Value)) %>%
        arrange(desc(avg))
```

4. What is the difference in the average levels of "EC PM2.5 LC TOR" between California and Arizona?

```{r}
filter(spec, State.Name %in% c("California", "Arizona") 
       & Parameter.Name == "EC PM2.5 LC TOR") %>%
       group_by(State.Name) %>%
        summarize(avg = mean(Sample.Value)) %>%
        spread(State.Name, avg) %>%
        mutate(diff = Arizona - California)
```

5. What are the median levels of "OC PM2.5 LC TOR" and "EC PM2.5 LC TOR" in the western and eastern U.S.? Define western as any monitoring location that has a `Longitude` less than -100.

```{r}
filter(spec, Parameter.Name %in% c("OC PM2.5 LC TOR", "EC PM2.5 LC TOR")) %>%
       mutate(region = ifelse(Longitude < -100, "west", "east")) %>%
        group_by(Parameter.Name, region) %>%
        summarize(median = median(Sample.Value)) %>%
        spread(region, median)
```


### Part 3

Use the `readxl` package to read the file
[aqs_sites.xlsx](../data/aqs_sites.xlsx) into R (you may need to
install the package first). You may get some warnings when reading in the data but you can ignore these for now.

```{r,warning=FALSE}
spec <- read_csv("data/SPEC_2014.csv.bz2", progress = FALSE)

library(readxl)
sites <- read_excel("data/aqs_sites.xlsx")
names(sites) <-  gsub(" +", ".", names(sites))
names(sites)
```

Use the functions in the `dplyr` and `tidyr` packages to answer the following questions.

1. How many monitoring sites are labelled as both "RESIDENTIAL" for `Land Use` and "SUBURBAN" for `Location Setting`?

```{r}
filter(sites, Land.Use == "RESIDENTIAL" & Location.Setting == "SUBURBAN") %>%
        summarize(n = n())
```

```{r}
with(sites, table(Land.Use, Location.Setting))
```

2. What are the median levels of "OC PM2.5 LC TOR" and "EC PM2.5 LC TOR" amongst monitoring sites that are labelled as both "RESIDENTIAL" and "SUBURBAN" in the eastern U.S., where eastern is defined as `Longitude` greater than or equal to -100?

```{r}
sites <- rename(sites, Site.Num = Site.Number) %>%
        select(State.Code, County.Code, Site.Num, Longitude, Land.Use, 
               Location.Setting)
spec <- mutate(spec, State.Code = as.numeric(State.Code),
               County.Code = as.numeric(County.Code),
               Site.Num = as.numeric(Site.Num)) %>%
        select(State.Code, County.Code, Site.Num, Parameter.Name, Sample.Value)
m <- left_join(spec, sites, by = c("State.Code", "County.Code", "Site.Num"))
str(m)
filter(m, Parameter.Name %in% c("OC PM2.5 LC TOR", "EC PM2.5 LC TOR")
       & Land.Use == "RESIDENTIAL" & Location.Setting == "SUBURBAN"
       & Longitude >= -100) %>%
        group_by(Parameter.Name) %>%
        summarize(median = median(Sample.Value))

```
